---
title: "Deaths Associated with Neurological Conditions"
author: "Evie Massheder"
date: "2024-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(scales)

library(dplyr)

library(openxlsx)

library(tidyr)

library(tidyverse)

library(here)

library(plotly)

```

## Project Description

Data is available from this link:

https://www.gov.uk/government/publications/deaths-associated-with-neurological-conditions

The URL for these pages is:

https://evie-lois.github.io/deaths-assoc-neuro/

The repository for these pages is:

https://github.com/evie-lois/deaths-assoc-neuro.git

## Research Question

Which neurological condition has the highest rate of associated deaths from 2001 to 2014?

## Import Data

```{r import}
# Importing one sheet from Excel workbook as dataset

# Specify the sheet by name from Excel workbook

data <- read.xlsx("/Users/eviemassheder/Documents/Data Analysis and Visualisation/workbook.xlsx", sheet = "Deaths in England") 

# Save the sheet to a new Excel file

write.xlsx(data, "/Users/eviemassheder/Documents/Data Analysis and Visualisation/DeathsinEngland.xlsx")

loc = "/Users/eviemassheder/Documents/Data Analysis and Visualisation/DeathsinEngland.xlsx"

data <- read.xlsx(loc)


```


## Data Preparations
```{r}
#Split embedded table into two

first_table <- data[1:16, ]

second_table <-data[17:nrow(data), ]

# Structure tables, exclude outliers 

first_table <- first_table[-c(16), ]

second_table <- second_table[-c(16:31),]

# Swap the first row of each table to the column names

colnames(first_table) <- first_table[1, ]

first_table <- first_table[-1, ]

colnames(second_table) <-second_table[1, ]

second_table <- second_table[-1, ]

# Clean data, removing outliers

data1_cleaned <- first_table[, -c(2,11)]

data2_cleaned <- second_table[, -c(8,10)]

# Combine tables

combined_table <- bind_cols(data1_cleaned, data2_cleaned)

# Clean new dataset

cleaned_dataset <- combined_table[,-c(10)]

#Code columns

cols <- c("year", "dwnc", "epi", "mndsma", "msid", "nmd", "poed", "tbsi", "totns", "at", "cnsi", "cnd", "dd", "fd", "ham", "raond", "smar")

colnames(cleaned_dataset) <- cols

#Reshape the data

reshaped_dataset <- cleaned_dataset %>%
  pivot_longer(cols = -year, 
               names_to = "Condition",
               values_to = "Deaths")

#Show cleaned and reshaped dataset

head(reshaped_dataset)

```

## Visualisation
```{r}
# Check 'Deaths' column is numeric for plot

str(reshaped_dataset$Deaths) 
reshaped_dataset$Deaths <- as.numeric(as.character(reshaped_dataset$Deaths))

# Create the scatter plot

ggplot(reshaped_dataset, aes(x = year, y = Deaths, color = Condition)) +
  geom_point(size = 3) +
  labs(title = "Deaths associated with Neurological Conditions (2001-2014)", 
       x = "Year", 
       y = "Number of Deaths") +
  theme_minimal() +
  scale_y_log10(labels = scales::comma) +
 
#Colour Code Conditions
  
  scale_color_manual(values = c( "dwnc" = "red", "epi" = "green", "mndsma" = "lightcoral", "msid" = "pink", "nmd"= "purple", "poed" = "orange", "tbsi" = "magenta", "totns" = "cyan", "at" = "blue", "cnsi" = "dodgerblue", "cnd" = "darkgreen", "dd" = "lightblue", "fd" = "violet", "ham" = "gold", "raond"= "darkblue", "smar" = "darkgray"), 
                     
 #Label Disorders
 
 labels = c( "dwnc" = "Deaths with Mention of Neurological Condition", "epi" = "Epliepsy", "mndsma" = "Motor Neurone Disease and Spinal Muscular Atrophy", "msid" = "Multiple Sclerosis and Inflammatory Disorders", "nmd" = "Neuromuscular Diseases", "poed" = "Parkinsonism and Other Etrapyrimidal Disorders", "tbsi" = "Traumatic brain and Spinal Injury", "totns" = "Tumours of the Nervous System", "at"= "Ataxia", "cnsi" = "CNS Infections", "cnd" = "Cranial Nerve Damage", "dd" = "Development Disorders", "fd" = "Functional Disorders", "ham" = "Headaches and Migraines", "raond" = "Rare and Other Neurological Diseases", "smar" = "Spondylotic Myelophathy and Radiculopathy")) +
  
# Flip the plot for horizontal view
  
  coord_flip()

#Interactive Plot 
  

```

## Summary

From completing this module and code project, I have learnt how to use R efficiently after having no previous knowledge of the software. I have also understood my mistakes and learned how to correct them. In addition to this...

If I was to do this project again with more time and data accessibility, I would have followed through with my previous choice of public health data looking into a comparison of the Western Diet and Mediterranean Diet. Comparing the effects on life expectancy and disease. I would have chosen data from individuals that follow either diet and compare their health to see if there was a correlation of disease and shorter life expectancy with the Western Diet to understand the effects of what we consume on our lives.
